%!TEX ROOT=ctutest.tex

\chapter{Introduction}

MDP's are one of the most well-known methods used for solving stochastic planning problems. In the last couple of years, a new programming language, Julia, has emerged into the scientific community's consciousness. While offering a high-level approach comparable to the one of Python, it also offers a similar speed of languages like C or C++ at the same time. As the new users come to this language, the demand for new tools rises. The subject of this work answers these calls and deals with implementing one of those methods, finite-horizon MDPs interface for Julia's POMDPs.jl package.

\section{Markov Decision Processes}
The Markov decision processes are a part of the optimization problem solvers.
The term MDP was probably first used in \cite{cite:1} and was derived from the name of Russian mathematician Andrey Markov who researched stochastic processes, on which are MDPs based \cite{wiki:1}.
The problems that are MDPs solving are described with the environment. The environment consists of the states and actions (with corresponding stochastic transitions) for which the agent either receives a reward or pays cost (negative reward). The agent's task is to maximize his reward. The problem can be finite or infinite according to the problem's description.

\section{Julia}
Julia is a young language that starts to receive traction for its features. According to https://star-history.t9t.io/#JuliaLang\julia the number of stars on its repository(https://github.com/JuliaLang/julia) tripled over the last 20 months and is rising steadily.
The development started in 2009 at MIT, and the project went public in 2012 [https://www.zdnet.com/article/is-julia-fastest-growing-new-programming-language-stats-chart-rapid-rise-in-2018/]. 
It combines the performance of low-level languages like Fortran or C with the high-level approach of Matlab, R and Python. In addition to that, Julia is dynamically typed, designed for parallelism and distributed computation, it offers multiple dispatch paradigms and is also reproducible while relying on community packages[https://julialang.org/].


\section{POMDPs.jl}
\textit{POMDPs.jl is a package providing a core interface for working with MDPs and POMDPs}\cite{JMLR:v18:16-300}x It offers the user a unified interface for custom problem definitions, problem interfaces for simpler ones, and multiple solvers for both MDPs and POMDPs. The package is managed by Zachary Sunberg and is under active development.


\section{Goal}
The goal of this project is to implement a finite horizon interface for MDPs in the POMDPs.jl package. To get to know how the package's environment works, develop possible test cases, solve them using a custom value iteration method, and then compare them with the already implemented method in POMDPs.jl. Devise a way to implement finite horizon for MDPs in the package and evaluate its correctness. Get feedback from the POMDPs.jl community, iterate the design, and submit a pull request to the corresponding repository.

