% arara: pdflatex: { synctex: yes }
% arara: makeindex: { style: ctuthesis }
% arara: bibtex

% The class takes all the key=value arguments that \ctusetup does,
% and a couple more: draft and oneside
\documentclass[twoside]{ctuthesis}

\DeclareUnicodeCharacter{2212}{-}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{float}

%\documentclass{article}
\renewenvironment{description}[1][0pt]
  {\list{}{\labelwidth=0pt \leftmargin=#1
   \let\makelabel\descriptionlabel}}
  {\endlist}

\parindent=0pt

\ctusetup{
% 	preprint = \ctuverlog,
	mainlanguage = english,
	titlelanguage = english,
% 	mainlanguage = czech,
% 	otherlanguages = {czech,english},
	title-czech = {MDP algoritmy v POMDPs.jl},
	title-english = {MDP algorithms in POMDPs.jl},
	subtitle-czech = {Zejména Finite-Horizon MDP},
	subtitle-english = {Finite-Horizon MDPs in particular},
	doctype = B,
	faculty = F3,
	department-czech = {Katedra Počítačů},
	department-english = {Department of Computer Science},
	author = {Tomáš Omasta},
	supervisor = {Ing. Jan Mrkos},
	supervisor-address = {E-325, \\ Charles Square 13, \\ Prague 2},
% 	supervisor-specialist = {John Doe},
	fieldofstudy-english = {Open Informatics},
	subfieldofstudy-english = {Artificial Intelligence and Computer Science},
	fieldofstudy-czech = {Otevřená Informatika},
	subfieldofstudy-czech = {Základy umělé inteligence a počítačových věd},
	keywords-czech = {Finite-Horizon, MDP, Julia},
	keywords-english = {Finite-Horizon, MDP, Julia},
	day = 10,
	month = 1,
	year = 2021,
	specification-file = {ctutest-zadani.pdf},
% 	front-specification = true,
	front-list-of-figures = false,
	front-list-of-tables = false,
%	monochrome = true,
%	layout-short = true,
}

\ctuprocess

\addto\ctucaptionsczech{%
	\def\supervisorname{Vedoucí}%
	\def\subfieldofstudyname{Studijní program}%
}

\ctutemplateset{maketitle twocolumn default}{
	\begin{twocolumnfrontmatterpage}
		\ctutemplate{twocolumn.thanks}
		\ctutemplate{twocolumn.declaration}
		\ctutemplate{twocolumn.abstract.in.titlelanguage}
		\ctutemplate{twocolumn.abstract.in.secondlanguage}
% 		\ctutemplate{twocolumn.tableofcontents}
% 		\ctutemplate{twocolumn.listoffigures}
	\end{twocolumnfrontmatterpage}
}

% Theorem declarations, this is the reasonable default, anybody can do what they wish.
% If you prefer theorems in italics rather than slanted, use \theoremstyle{plainit}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{note}
\newtheorem*{remark*}{Remark}
\newtheorem{remark}[theorem]{Remark}

\setlength{\parskip}{5ex plus 0.2ex minus 0.2ex}

% Abstract in Czech
\begin{abstract-czech}
Do povědomí vědecké komunity se začíná dostávat nový programovací jazyk, Julia. Díky svému prudkému nárustu uživatelské báze a svému prostředí postavenému na open-source knihovnách, se dá očekávat, že se stane jedním z nejpopulárnějších jazyků. V této práci představíme Julii, MDP, teorii s nimi spjatou spolu s možnými přístupy k jejich řešení. Postupně ukážeme jejich výhody i nevýhody a nakonec se dostaneme k Finite-Horizon MDP, jako jedné z nejpokročilejších metod. Navrhneme design Finite-Horizon MDP rozhraní na základě koncepce z JuliaPOMDP/FiniteHorizonPOMDPs.jl \cite{FHPOMDP}, implementujeme jej a porovnáme s value iteration řešením DiscreteValueIteration.jl \cite{DVI} implementovaným v knihovně JuliaPOMDP's \cite{JuliaPOMDP}.
\end{abstract-czech}

% Abstract in English
\begin{abstract-english}
In the last couple of years, a new programming language, Julia, has emerged into the scientific community's consciousness. With the steep rise of its downloads and its environment built on open-source packages, it is expected to become one of the most popular languages. This project introduces Julia, MDPs, the theory behind it, and possible approaches to its solution. By reviewing its drawbacks and advantages, we gradually get to present Finite-Horizon MDPs as the advanced solution. We design Finite-Horizon MDP Interface according to declaration at JuliaPOMDP/FiniteHorizonPOMDPs.jl \cite{FHPOMDP}, implement the interface and benchmark it against JuliaPOMDP's \cite{JuliaPOMDP} value iterator solver DiscreteValueIteration.jl \cite{DVI}.

\end{abstract-english}

% Acknowledgements / Podekovani
\begin{thanks}
We thank our Supervisor Ing. Jan Mrkos for all the little things he has done for us in making this possible, particularly for his patience. Our thanks also belong to our family and our girlfriend for the support they provided us with, and finally, to FEE CTU for all joys and sorrows it enriched us with.
\end{thanks}

% Declaration / Prohlaseni
\begin{declaration}
I declare that this work is all my own work and I have cited all sources I have
used in the bibliography.

\medskip

Prague, \monthinlanguage{title} \ctufield{day}, \ctufield{year}

\vspace*{2cm}

Prohlašuji, že jsem předloženou práci vypracoval samostatně, a že jsem uvedl veškerou použitou literaturu.

\medskip

V Praze, \ctufield{day}.~\monthinlanguage{second}~\ctufield{year}
\end{declaration}

% Only for testing purposes
\listfiles
\usepackage[pagewise]{lineno}
\usepackage{lipsum,blindtext}
\usepackage{mathrsfs} % provides \mathscr used in the ridiculous examples

\begin{document}

\maketitle

\input{ctutest-Intro}

\input{ctutest-MDP-Theory}

\input{ctutest-MDP-Implementation}

\input{ctutest-POMDP-Theory}

\input{ctutest-POMDP-Implementation}

\input{ctutest-EX-problems}

\input{ctutest-EX-benchmark}

\input{ctutest-Validation}

\input{ctutest-Profiling}

\input{ctutest-Conclusion}

\appendix

% \printindex


\bibliographystyle{amsalpha}
\bibliography{ctutest}

% \ctutemplate{specification.as.chapter}

\end{document}